Variance reduction for MCMC methods via martingale representations

### 2020-02-19 aap/Revision/vr-mcmc-revision-new.tex
update

f(X^x_{p})=\mathsf{E}\left[\left.f(X^x_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}a_{p,l,k}(X^x_{l-1})\phi_k\left(\xi_{l}\right),

@@

\begin{proof}%[Proof of Theorem~\ref{prop:29032018a1}]
The expansion obviously holds for $p=1$ and $j=0$.
Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system in \(L^2(\mathbb{R}^d, P_{\xi})\), we have for all $x \in \rset^d$,
\eric{it is clear, but we must say in which sense this equality should be understood}
\[
f(X^x_{1})=\PE[f(X_{1}^x)]+\sum_{k\geq1}a_{1,1,k}(x)\phi_{k}(\xi_{1})
\]
with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$,
provided $\PE[|f(X^x_{1})|^{2}]<\infty.$
\tcr{Recall that $\mathcal{G}_l=\sigma(\xi_{1},\ldots,\xi_{l}),$
$l=1,2,\ldots, $
and $\mathcal{G}_0=\mathrm{triv}$.}
\eric{do we need this, this is recalled ten lines above}
Suppose that (\ref{eq:mart_repr})
holds for $p=q$, all $j<q$, and all Borel-measurable functions
$f$ with $\mathsf{E}\left[|f(X_{q})|^2\right]<\infty$.
Let us prove it for $p=q+1$.
Given $f$ with $\mathsf{E}\left[|f(X_{p})|^2\right]<\infty$,
due to the orthonormality and completeness
of the system $\left(\phi_{k}\right)$, we get by
conditioning on $\mathcal{G}_{q}$,
\[
f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|{\mathcal{G}}_q\right]+\sum_{k\geq1}\alpha_{p,q+1,k}\phi_{k}(\xi_{q+1}),
\]
where
\begin{equation*}
\alpha_{p,q+1,k}
=\mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{q+1})\right|\mathcal G_q\right].
\end{equation*}
By the Markov property of $(X_{l})$,
we have
$\mathsf{E}[f(X_{p})|\mathcal{G}_q]
=\mathsf{E}[f(X_{p})|X_{q}]$.
Furthermore, a calculation involving
intermediate conditioning on $\mathcal G_{q+1}$
and the recurrence relation
$X_{q+1}=\Phi_{q+1}(X_{q},\xi_{q+1})$
verifies that
\[
\alpha_{p,q+1,k}=\mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{q+1})\right|X_{q}\right]
=a_{p,q+1,k}(X_{q})
\]
for suitably chosen
Borel-measurable functions $a_{p,q+1,k}$.
We thus arrive at
\begin{align}
\label{berm:sig_X}
f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|X_{q}\right]+\sum_{k\geq1}a_{p,q+1,k}(X_{q})\phi_{k}(\xi_{q+1}),
\end{align}
which is the required statement in the case $j=q$.
Now assume $j<q$.
The random variable
$\mathsf{E}\left[\left.f(X_{p})\right|X_{q}\right]$
is square integrable and has the form
$g(X_{q})$,
hence the induction hypothesis applies, and we get
\begin{equation}\label{eq:28082017a1}
\mathsf{E}\left[\left.f(X_{p})\right|X_{q}\right]=\mathsf{E}\left[\left.f(X_{p})\right|X_{j}\right]+\sum_{k\geq1}\sum_{l=j+1}^{q}a_{p,l,k}(X_{l-1})\phi_{k}(\xi_{l})
\end{equation}
with
\begin{align*}
a_{p,l,k}(X_{l-1}) &= \mathsf{E}\left[\left.\mathsf{E}\left[\left.f(X_{p})\right|\mathcal{G}_{q}\right]\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]
= \mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]\\
&=\mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{l})\right|X_{l-1}\right].
\end{align*}
Formulas \eqref{berm:sig_X}
and~\eqref{eq:28082017a1} conclude the proof.
\end{proof}

@@

\section{Proofs}
\label{sec:proofs}
\subsection{Proof of Theorem~\ref{th:mr}}
For $l\le p$ and $x \in \rset^d$, we have the representation

@@

X^x_{p}=G_{p}(x,\sqrt{\gamma}Z_{1},\ldots,\sqrt{\gamma}Z_{p}),

@@

where the function \(G_{p}:\) \(\mathbb{R}^{d\times(p+1)}\to \mathbb{R}^{d}\) is defined as
G_{p}(x,y_1,\ldots,y_p):=\Phi(\cdot,y_{p})\circ\Phi(\cdot,y_{p-1})\circ\ldots\circ\Phi(x,y_{1})
with, for $x,y\in\mathbb R^d$,   $\Phi(x,y)=x-\gamma\mu(x)+y$. 
As a consequence, for a function $f\colon\mathbb R^d\to\mathbb R$
as in Section~\ref{sec:setup}, we have
f\left(X_{p}\right) =f\circ G_{p}(X_{0},\sqrt{\gamma}Z_{1},\ldots,\sqrt{\gamma}Z_{p}).
In what follows, for $\mathbf{k}\in\mathbb N_0^d$,
we use the shorthand notation
\label{eq:definition-differential-f-p}
\partial_{1}^{\mathbf k} f\left(X_{p}\right)
:=\partial_{1}^{\mathbf k} [f\circ G_{p}](X_{0},\sqrt{\gamma}Z_{1},\ldots,\sqrt{\gamma}Z_{p})
whenever the function $f\circ G_{p}$ is smooth enough
(that is, $f$ and $\mu$ need to be smooth enough).
Finally, for a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$, we use the notation
\(\mathbf{k} ! :=k_1!\cdot\ldots\cdot k_d !\)

>>>>

f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}a_{p,l,k}(X_{l-1})\phi_k\left(\xi_{l}\right),

@@

\begin{proof}%[Proof of Theorem~\ref{prop:29032018a1}]
The expansion obviously holds for $p=1$ and $j=0$.
Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system in \(L^2(\mathbb{R}^d, P_{\xi})\), we have for all $x \in \rset^d$,
\eric{it is clear, but we must say in which sense this equality should be understood}
\[
f(X_{1})=\PE[f(X_{1})]+\sum_{k\geq1}a_{1,1,k}(x)\phi_{k}(\xi_{1})
\]
with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$,
provided $\PE[|f(X_{1})|^{2}]<\infty.$
\tcr{Recall that $\mathcal{G}_l=\sigma(\xi_{1},\ldots,\xi_{l}),$
$l=1,2,\ldots, $
and $\mathcal{G}_0=\mathrm{triv}$.}
\eric{do we need this, this is recalled ten lines above}
Suppose that (\ref{eq:mart_repr})
holds for $p=q$, all $j<q$, and all Borel-measurable functions
$f$ with $\mathsf{E}\left[|f(X_{q})|^2\right]<\infty$.
Let us prove it for $p=q+1$.
Given $f$ with $\mathsf{E}\left[|f(X_{p})|^2\right]<\infty$,
due to the orthonormality and completeness
of the system $\left(\phi_{k}\right)$, we get by
conditioning on $\mathcal{G}_{q}$,
\[
f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|{\mathcal{G}}_q\right]+\sum_{k\geq1}\alpha_{p,q+1,k}\phi_{k}(\xi_{q+1}),
\]
where
\begin{equation*}
\alpha_{p,q+1,k}
=\mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{q+1})\right|\mathcal G_q\right].
\end{equation*}
By the Markov property of $(X_{l})$,
we have
$\mathsf{E}[f(X_{p})|\mathcal{G}_q]
=\mathsf{E}[f(X_{p})|X_{q}]$.
Furthermore, a calculation involving
intermediate conditioning on $\mathcal G_{q+1}$
and the recurrence relation
$X_{q+1}=\Phi_{q+1}(X_{q},\xi_{q+1})$
verifies that
\[
\alpha_{p,q+1,k}=\mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{q+1})\right|X_{q}\right]
=a_{p,q+1,k}(X_{q})
\]
for suitably chosen
Borel-measurable functions $a_{p,q+1,k}$.
We thus arrive at
\begin{align}
\label{berm:sig_X}
f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|X_{q}\right]+\sum_{k\geq1}a_{p,q+1,k}(X_{q})\phi_{k}(\xi_{q+1}),
\end{align}
which is the required statement in the case $j=q$.
Now assume $j<q$.
The random variable
$\mathsf{E}\left[\left.f(X_{p})\right|X_{q}\right]$
is square integrable and has the form
$g(X_{q})$,
hence the induction hypothesis applies, and we get
\begin{equation}\label{eq:28082017a1}
\mathsf{E}\left[\left.f(X_{p})\right|X_{q}\right]=\mathsf{E}\left[\left.f(X_{p})\right|X_{j}\right]+\sum_{k\geq1}\sum_{l=j+1}^{q}a_{p,l,k}(X_{l-1})\phi_{k}(\xi_{l})
\end{equation}
with
\begin{align*}
a_{p,l,k}(X_{l-1}) &= \mathsf{E}\left[\left.\mathsf{E}\left[\left.f(X_{p})\right|\mathcal{G}_{q}\right]\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]
= \mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]\\
&=\mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{l})\right|X_{l-1}\right].
\end{align*}
Formulas \eqref{berm:sig_X}
and~\eqref{eq:28082017a1} conclude the proof.
\end{proof}

@@

\section{Numerical results}
\label{sec:num}
If each function  \(Q_{p,l}(x) \) can be well approximated by polynomials, then we can use polynomial basis functions in   \eqref{eq:06042018a1} to approximate \(\bar Q_l\). Suppose that we constructed a polynomial approximation  for each function \(\bar Q_{l}\)  in the form:
\widehat{Q}_{l}(x) = \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} x^{\mathbf{s}},\quad s=(s_1,\ldots,s_d)
with some coefficients \(\beta_{\mathbf{s}}\in \mathbb{R}.\)
Then using the identity

@@

\xi^j = j! \sum_{r = 0}^{\lfloor j/2 \rfloor} \frac{1}{2^r  r! \sqrt{(j-2r)!}} H_{j-2r}(\xi),\quad \xi \in \mathbb{R},

@@

we derive for all $x \in \mathbb{R}^d$,
\widehat a_{l,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(x) \widehat{Q}_{l}(x - \gamma_l \mu(x) + \sqrt{\gamma_l}\xi)}{\TrainSet \vee \mathcal{G}_N}
= \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{i,k_i,s_i}(x),
where for all integers $i,k,s_i$ and $x \in \rset^d$,
P_{l,k,s}(x) &= \mathsf{E} \left[ H_{k}(\xi_l) (x_l - \gamma_l \mu_l(x) + \sqrt{\gamma_l} \xi_l)^{s} \right]
is a one-dimensional polynomial (in \(x\)) of degree at most \(s\) with analytically known coefficients. Alternatively, on practice we can work directly with functions \(Q_{p,l}.\) Due to (almost) time-homogenous Markov chain generated by ULA, each function $Q_{p,l}$ can be approximated by function, which depends only on the  $p-l$, that is,
 Q_{p,l}(x) = \mathsf{E} \left[f(X_p) | X_l = x\right] \approx Q^\circ_{p-l}(x)
Thus we have
  Q^\circ_{r}(x) = \mathsf{E} \left[f(X_{l+r}) | X_l = x \right].
Consequently, the functions $Q^\circ_r$ can be estimated using a modified least squares criteria
\label{eq:Qcircr}
\widehat{Q}^\circ_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2
for \(1 \leq r \leq n-1,\) where  $\widehat{Q}^\circ_0(x) = f(x)$ by definition.
Due to Lemma~\ref{a_decay}, it is enough to estimate $Q^\circ_{r}$ for $|r| < n_{\mathrm{trunc}}$ for some truncation level \(n_{\mathrm{trunc}}\) depending on \(d\) and \(\gamma.\) It allows us to use a smaller amount of training trajectories to approximate conditional expectations $Q^\circ_{r}$.
Finally  we construct a truncated version of our estimator:
\pi_{K,n,n_{\mathrm{trunc}}}^N(f) = \pi_n^N(f) -  M_{K,n, n_{\mathrm{trunc}}}^{N}(f),
{\widehat  M}_{K,n, n_{\mathrm{trunc}}}^N &=& \sum_{0<\left\| \mathbf{k}\right\| \leq K} \sum_{p=N+1}^{N+n} \omega_{p,n}^N
&&\times
\sum_{l=N+1}^p {\widehat  a}_{p-l,\mathbf{k}}(X_{l-1})\mathbf{H}_{\mathbf{k}}(Z_l) \mathbbm{1}\{ |p-l| < n_{\mathrm{trunc}} \}
and
\widehat a_{p-l,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(x) \widehat{Q}^\circ_{p-l}(x - \gamma_l \mu(x) + \sqrt{\gamma_l}\xi)}{\TrainSet \vee \mathcal{G}_N}.
\subsection{Gaussian mixtures}
 We consider a sample generated by ULA with $\pi$  given by the mixture of two Gaussian distributions  with equal weights:
\pi (x) = \frac{1}{2(2\pi)^{d/2}} \left( \rme^{\frac{-\| x-a\|_2^2}{2}}  + \rme^{\frac{-\| x+a\|_2^2}{2}} \right), \quad x \in\mathbb R^d
where $a \in \mathbb{R}^d$ is a given vector. The function $U(x)$ and its gradient are given by
U(x) = \frac{1}{2} \|x - a\|_2^2 - \text{log}(1 + \rme^{-2x^\top a}), \,\,
\nabla U(x) = x-a +2a(1 + \rme^{2 x^\top a})^{-1},
respectively.
In our experiments we considered dimensions $d = 2$ and $d = 8$ and take $a=( (2d)^{-1/2}, \dots, (2d)^{-1/2})^\top$.
In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i,\)
 we have used constant step sizes $\gamma_i = \gamma=0.2$ and sampled $n = 10^4$ samples for $d = 2$ and $n = 2 \times 10^3$ for $d = 8$.  We generated $\NtrainPath = 10$ independent "training" trajectories and solved the least squares  problems \eqref{eq:Qcircr} using the first order polynomial approximations for the coefficients \(a_{p,\mathbf{k}}\) as described in the previous section. The truncation level \(n_{\mathrm{trunc}}\) is chosen to be \(50.\) To test our variance reduction algorithm, we generated 100 independent trajectories.
 In Figure ~\ref{fig:1} we compare our approach to variance reduction methods of \cite{mira2013zero} and \cite{brosse2018diffusion}.

\section{Proofs}
\label{sec:proofs}
\subsection{Proof of Theorem~\ref{th:mr}}
For $l\le p$ and $x \in \rset^d$, we have the representation
\[
X^x_{p}=G_{p}(x,\sqrt{\gamma}Z_{1},\ldots,\sqrt{\gamma}Z_{p}),
\]
where the function \(G_{p}:\) \(\mathbb{R}^{d\times(p+1)}\to \mathbb{R}^{d}\) is defined as
\begin{equation}
\label{eq:definition-G-p-l}
G_{p}(x,y_1,\ldots,y_p):=\Phi(\cdot,y_{p})\circ\Phi(\cdot,y_{p-1})\circ\ldots\circ\Phi(x,y_{1})
\end{equation}
with, for $x,y\in\mathbb R^d$,   $\Phi(x,y)=x-\gamma\mu(x)+y$. 
As a consequence, for a function $f\colon\mathbb R^d\to\mathbb R$
as in Section~\ref{sec:setup}, we have
$$
f\left(X_{p}\right) =f\circ G_{p}(X_{0},\sqrt{\gamma}Z_{1},\ldots,\sqrt{\gamma}Z_{p}).
$$
In what follows, for $\mathbf{k}\in\mathbb N_0^d$,
we use the shorthand notation
\begin{equation}
\label{eq:definition-differential-f-p}
\partial_{1}^{\mathbf k} f\left(X_{p}\right)
:=\partial_{1}^{\mathbf k} [f\circ G_{p}](X_{0},\sqrt{\gamma}Z_{1},\ldots,\sqrt{\gamma}Z_{p})
\end{equation}
whenever the function $f\circ G_{p}$ is smooth enough
(that is, $f$ and $\mu$ need to be smooth enough).
Finally, for a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$, we use the notation
\(\mathbf{k} ! :=k_1!\cdot\ldots\cdot k_d !\)
